A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-05-12 paper/paper.Rmd
minor typos

<!--AE comment: first sentence can be removed or rephrased. Next sentence includes the word wages too many times. Next sentence should say “dropouts” rather than “drop outs” to be consistent with the previous sentence. -->
The original set, containing wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994, is available in the R package `brolgar`. To compare the refreshed data with the original, a subset needs to be matched. There are numerous ways to do this, with the simplest being to extract the individuals based on their id being part of the original data and restricting the longitudinal measurements to the same years. However, we decided to try to replicate the process, as suggested by the description of the original data. This requires first identifying individuals who dropped out of high school.

@@

The wages data provides a good opportunity to discuss the difference between statistics for public policy and statistics that relate to the individual. Public policy is based on models, yielding averages that might vary across strata. Modeling the wages relative to workforce experience with demographic covariates, we learn that there are significantly different patterns. More education leads to increasingly higher wages, which is a satisfying result for educators. It means education makes a difference in the wage experience and that public policy that encourages education is a data-supported action. We would also learn, although not presented here, that wages on average differ by race category. This is a disturbing finding because there is no rationale for such a difference in a fair society. This would provide support for action in public policy to remove this overall average effect. It also provides an example for educators to explain the use of data to support public policy action.

@@

- **R Package**:`r yowie` is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort. This package can be accessed from \url{`r yowie_repo`}.

>>>>

<!--AE comment: first sentence can be removed or rephrased. Next sentence includes the word wages too many times. Next sentence should say “dropouts” rather than “drop outs” to be consistent with the previous sentence. -->
The original set, containing wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994, is available in the R package `brolgar` [@brolgar]. To compare the refreshed data with the original, a subset needs to be matched. There are numerous ways to do this, with the simplest being to extract the individuals based on their id being part of the original data and restricting the longitudinal measurements to the same years. However, we decided to try to replicate the process, as suggested by the description of the original data. This requires first identifying individuals who dropped out of high school.

@@

The wages data provides a good opportunity to discuss the difference between statistics for public policy and statistics that relate to the individual. Public policy is based on models, yielding averages that might vary across strata. Modeling the wages relative to workforce experience with demographic covariates, we learn that there are significantly different patterns. More education leads to increasingly higher wages, which is a satisfying result for educators. It means education makes a difference in the wage experience, so public policy encouraging education is a data-supported action. We would also learn, although not presented here, that wages on average differ by race category. This is a disturbing finding because there is no rationale for such a difference in a fair society. This would provide support for action in public policy to remove this overall average effect. It also provides an example for educators to explain the use of data to support public policy action.

@@

- **R Package**: `r yowie` is an R data package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort. This package can be accessed from \url{`r yowie_repo`}.
